---
title: 'Unit 1 Case Study: Hashtag Common Core'
subtitle: "ECI 589 Intro to SNA and Education"
author: "Dr. Shaun Kellogg"
date: "`r format(Sys.Date(),'%B %e, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
    code_folding: show
    code_download: TRUE
editor_options:
  markdown:
    wrap: 72
bibliography: lit/references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

For this independent practice, you will dive into and interpret the fit measures in more depth.

## Overview of classification measures

- We are interested in predicting a _dichotomous_ or _continuous_ outcome; our predictions are _probabilities_
- These probabilities are often changed (through the use of a threshold - often 0.50) back into dichotomous or categorical labels
- Metrics for how well the model performed include:
  - *Accuracy*: For the known codes, what percentage of the predictions are correct
  - *Cohen's $\kappa$*: Same as accuracy, while account for the base rate of (chance) agreement
  - *Sensitivity (AKA recall)*: Among the true "positives", what percentage are classified as "positive"?
  - *Specificity*: Among the true "negatives", what percentage are classified as "negative"?
  - *ROC AUC*: For different levels of the threshold, what is the sensitivity and specificity?

```{r}

```