---
  title: "SNA Lab 4: Statistical Inference & Network Models"
  subtitle: "Test"
  author: "**Dr. Shaun Kellogg**"
  institute: "LASER Institute"
  date: '`r format(Sys.time(), "%B %d, %Y")`'
  output:
    xaringan::moon_reader:
      css:
       - default
       - css/laser.css
       - css/laser-fonts.css
      lib_dir: libs                        # creates directory for libraries
      seal: false                          # false: custom title slide
      nature:
        highlightStyle: default         # highlighting syntax for code
        highlightLines: true               # true: enables code line highlighting 
        highlightLanguage: ["r"]           # languages to highlight
        countIncrementalSlides: false      # false: disables counting of incremental slides
        ratio: "16:9"                      # 4:3 for standard size,16:9
        slideNumberFormat: |
         <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
         </div>
---
class: clear, title-slide, inverse, center, top, middle

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=FALSE}
# then load all the relevant packages
pacman::p_load(pacman, knitr, tidyverse, readxl)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

```{r xaringanExtra-clipboard, echo=FALSE}
# these allow any code snippets to be copied to the clipboard so they 
# can be pasted easily
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
  ),
  rmarkdown::html_dependency_font_awesome()
)
```
```{r xaringan-extras, echo=FALSE}
xaringanExtra::use_tile_view()

```

# `r rmarkdown::metadata$title`
----
### `r rmarkdown::metadata$author`
### `r format(Sys.time(), "%B %d, %Y")`

---
# Agenda

.pull-left[
## Part 1: Core Concepts
### Data Management
- xx
- xx
- xx
]

.pull-right[

## Part 2: R Code-Along
#### Exponential Random Graph Models
- xx
- xx
- xx
]


.panelset[
.panel[.panel-name[1]

---

# How do I select a model?

One general principle is to **start with the simplest useful model** and to _build toward
more complex models as helpfuL_.

This principle applies in multiple ways:

- To choose an algorithm, start with simpler models that you can efficiently use and understand
- To carry out feature engineering, understand your predictors well by starting with a subset
- To tune an algorithm, start with a relatively simple set of tuning parameters

This isn't just for beginners or those of us in education; [most spam filters use Support Vector Machines (and used Naive Bayes until recently)](https://vas3k.com/blog/machine_learning/) due to their combination of effectiveness and efficiency "in production."

---

- Nothing too much, apart from computing power, time, and concerns of 
- A "check" on your work is your predictions on _test_ set data
  - *Train data*: Coded/outcome data that you use to train ("estimate") your model
  - *Validation data<sup>1</sup>*: Data you use to select a particular algorithm
  - *Test ("hold-out") data*: Data that you do not use in any way to train your data

- An important way to achieve good performance with test data is to balance between the inherent _bias_ in your algorithm and the _variance_ in the predictions of your algorithm; this is referred to as the **bias-variance** trade-off of _all_ models

.footnote[
[1] not always/often used, for reasons we'll discuss later

]

---

# Overview of regression modeling

1. Split data (into train, test, and tuning sets sets)
1. Specify recipe, model, and workflow
1. Estimate models and determine tuning parameters' values
1. Evaluate metrics and predictions

---

# Data from online science classes

- This data comes from a study of ~700 secondary/high school-aged students in the United States
- These were "one-off" classes that helped to fill gaps in students' schedules
- The data were collected _over multiple semesters_ from _multiple classes_
- There are a number of types of variables:
  - Demographic/contextual variables, e.g. `subject` and `gender`
  - Self-report survey variables: `uv` (utility value), `percomp` (perceived competence), and `int` (interest)
  - Gradebook variables: `percentage_earned` (based on the first 20 assignments)
  - Discussion variables: `sum_discussion_posts`, `sum_n_words` (for the first 3 discussions)
  - Outcomes: `final_grade`, `passing_grade`, `time_spent` (in minutes)

---

# Let's look at these steps once more

1. Split data (into train, test, and tuning sets sets)
1. Specify recipe, model, and workflow
1. Estimate models and determine tuning parameters' values
1. Evaluate metrics and predictions

---

class: clear, center

## .font130[.center[**Thank you!**]]
<br/>
.center[<img style="border-radius: 80%;" src="img/jr-cycling" height="200px"/><br/>**Dr. Joshua Rosrenberg**<br/><mailto:jmrosenberg@utk.edu>]
