---
  title: "SNA Lab 4: Statistical Inference & Network Models"
  subtitle: "Test"
  author: "**Dr. Shaun Kellogg**"
  institute: "LASER Institute"
  date: '`r format(Sys.time(), "%B %d, %Y")`'
  output:
    xaringan::moon_reader:
      css:
       - default
       - css/laser.css
       - css/laser-fonts.css
      lib_dir: libs                        # creates directory for libraries
      seal: false                          # false: custom title slide
      nature:
        highlightStyle: default         # highlighting syntax for code
        highlightLines: true               # true: enables code line highlighting 
        highlightLanguage: ["r"]           # languages to highlight
        countIncrementalSlides: false      # false: disables counting of incremental slides
        ratio: "16:9"                      # 4:3 for standard size,16:9
        slideNumberFormat: |
         <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
         </div>
---
class: clear, title-slide, inverse, center, top, middle

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=FALSE}
# then load all the relevant packages
pacman::p_load(pacman, knitr, tidyverse, readxl)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

```{r xaringanExtra-clipboard, echo=FALSE}
# these allow any code snippets to be copied to the clipboard so they 
# can be pasted easily
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
  ),
  rmarkdown::html_dependency_font_awesome()
)
```
```{r xaringan-extras, echo=FALSE}
xaringanExtra::use_tile_view()

```

# `r rmarkdown::metadata$title`
----
### `r rmarkdown::metadata$author`
### `r format(Sys.time(), "%B %d, %Y")`

---
# Agenda

.pull-left[
## Part 1: Core Concepts
### Data Management
- xx
- xx
- xx
]

.pull-right[

## Part 2: R Code-Along
#### Exponential Random Graph Models
- xx
- xx
- xx
]


.panelset[
.panel[.panel-name[1]

---

---

# Overview of regression modeling

1. Split data (into train, test, and tuning sets sets)
1. Specify recipe, model, and workflow
1. Estimate models and determine tuning parameters' values
1. Evaluate metrics and predictions

---

# Data from online science classes

- This data comes from a study of ~700 secondary/high school-aged students in the United States
- These were "one-off" classes that helped to fill gaps in students' schedules
- The data were collected _over multiple semesters_ from _multiple classes_
- There are a number of types of variables:
  - Demographic/contextual variables, e.g. `subject` and `gender`
  - Self-report survey variables: `uv` (utility value), `percomp` (perceived competence), and `int` (interest)
  - Gradebook variables: `percentage_earned` (based on the first 20 assignments)
  - Discussion variables: `sum_discussion_posts`, `sum_n_words` (for the first 3 discussions)
  - Outcomes: `final_grade`, `passing_grade`, `time_spent` (in minutes)

---

# Let's look at these steps once more

1. Split data (into train, test, and tuning sets sets)
1. Specify recipe, model, and workflow
1. Estimate models and determine tuning parameters' values
1. Evaluate metrics and predictions

---

class: clear, center

## .font130[.center[**Thank you!**]]
<br/>
.center[<img style="border-radius: 80%;" src="img/jr-cycling" height="200px"/><br/>**Dr. Joshua Rosrenberg**<br/><mailto:jmrosenberg@utk.edu>]
