---
title: 'Learning Lab 2 Guided Practice'
author: ""
date: "`r format(Sys.Date(),'%B %e, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
    code_folding: show
    code_download: TRUE
editor_options:
  markdown:
    wrap: 72
bibliography: lit/references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This case study is similar to the first, but it differs in two key ways:

1. We use a very different data set, on from _online science classes_ that involves a variety of variables types
2. We focus on _feature engineering_, a key step in which we prepare variables for inclusion in our machine learning models

Studying online classes has been a longstanding focus of the learning analytics community. Consider this paper by Li et al. (2020) published in *The Internet and Higher Education* available here: https://github.com/laser-institute/essential-readings/blob/main/machine-learning/ml-lab-2/li-et-al-2020-ihe.pdf

In this study, digital _log-trace data_, data generated through users' interactions with digital technologies, was used to study elements of the theoretical frame of _self-regulated learning_ and how it related to students' achievement. We use similar data in this study, treating students' interactions with the online course learning management system - and other variables, including self-report measures, to predict learners' achievement. This case study is related to this one [here](https://datascienceineducation.com/c14.html) in the _Data Science in Education Using R_ book.

## Step 0: Loading and setting up

Like in the first learning lab, we'll load several packages, first.

```{r, load packages}
library(tidyverse)
library(here)
library(tidymodels)
library(vip)
library(ranger)
```

# Let's take a look at the data

```{r, include = FALSE}
d <- read_csv("data/data-to-model.csv")

d <- select(d, -time_spent) # this is another outcome

d <- mutate(d, passing_grade = ifelse(final_grade > 70, 1, 0),
                  passing_grade = as.factor(passing_grade)) %>% 
    select(-final_grade)
```

```{r}
d %>% 
    glimpse()
```

## Step 1. Split data

Next, we'll split the data, just like before. We use an 80% split again; how will you "spend" your data? You can change this number if you wish, but cnsider how much data you have to "spend" for both training and testing.

```{r}
train_test_split <- initial_split(d_class, prop = .80)

data_train <- training(train_test_split)
```

Here's a key difference! Pay careful attention to this next line of code

```{r}
kfcv <- vfold_cv(data_train) # this differentiates this from what we did before
# before, we simple used data_train to fit our model
```

## Step 2: Engineer features

We'll engage in feature engineering, like before, but, here, we add considerably more steps.

#### [Your Turn]{style="color: green;"} ⤵ {style="font-style: normal; font-variant-caps: normal; letter-spacing: normal; orphans: auto; text-align: start; text-indent: 0px; text-transform: none; white-space: normal; widows: auto; word-spacing: 0px; -webkit-tap-highlight-color: rgba(26, 26, 26, 0.3); -webkit-text-size-adjust: auto; -webkit-text-stroke-width: 0px; text-decoration: none; caret-color: rgb(0, 0, 0); color: rgb(0, 0, 0);"}

Read about (here: https://www.tmwr.org/recipes.html) and add the following four steps to the code chunk below.

- add_role(student_id, course_id, new_role = "ID variable")  
- step_novel(all_nominal_predictors())  
- step_dummy(all_nominal_predictors())  
- step_impute_knn(all_predictors(), all_outcomes())  
    
```{r panel-chunk-2, echo = TRUE, eval = FALSE}
sci_rec <- recipe(passing_grade ~ ., data = d_class) %>% 
    step_normalize(all_numeric_predictors()) %>%
    step_nzv(all_predictors())
```

## Step 3: Specify recipe, model, and workflow

Here, we specify the same logistic regression as above:

- using the `logistic_reg()` function to set the *model*
- using `set_engine("glm")` to set the *engine*
- finally, using `set_mode("classification"))` to set the "*mode* to classification; this could be changed to regression for a continuous/numeric outcome
 
```{r panel-chunk-3, echo = TRUE, eval = FALSE}
# specify model
rf_mod_many <-
    rand_forest() %>%
    set_engine("ranger", importance = "impurity") %>%
    set_mode("classification")
```

Last, we'll put the pieces together - the model and recipe.

```{r}
# specify workflow
rf_wf_many <-
    workflow() %>%
    add_model(rf_mod_many) %>% 
    add_recipe(sci_rec)
```
]

## Step 4: Fit model

Note that here we use the `kfcv` data. We'll run that in the next chunk.

```{r panel-chunk-4, echo = TRUE, eval = FALSE}
tree_res <- fit_resamples(rf_wf_many, data = kfcv)
```

How does this output differ from that from fitting the data on a single (training) set, as in the first learning lab? Add one or more ideas here:

- 

Finally, we'll use the `last_fit` function to fit the model using the entire data set.

```{r}
final_fit <- last_fit(tree_res, train_test_split,
               metrics = metric_set(roc_auc, accuracy, kap, 
                                                    sensitivity, specificity, precision))
```

## Step 5: Interpret accuracy

```{r panel-chunk-5, echo = TRUE, eval = FALSE}
# fit stats
final_fit %>%
    collect_metrics()
```

Run the code below to calculate a _confusion matrix_:

```{r}
final_fit$.predictions[[1]] %>% 
    conf_mat(.pred_class, code)
```

What do you notice from the above two? More generally, how did the model do? 

- 

Considering what you've read in the two required readings, is this suitable and appropriate for scaling up? Add several notes on this, to

-

### 🧶 Knit & Check ✅

Congratulations - you've completed the Machine Learning Learning Lab 1 Guided Practice! Consider moving on to the independent practice next.