---
title: "Machine Learning Learning Lab 2"
subtitle: "Overview Presentation"
author: "**Dr. Joshua Rosenberg**"
institute: "LASER Institute"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  xaringan::moon_reader:
    css:
     - default
     - css/laser.css
     - css/laser-fonts.css
    lib_dir: libs                        # creates directory for libraries
    seal: false                          # false: custom title slide
    nature:
      highlightStyle: default         # highlighting syntax for code
      highlightLines: true               # true: enables code line highlighting 
      highlightLanguage: ["r"]           # languages to highlight
      countIncrementalSlides: false      # false: disables counting of incremental slides
      ratio: "16:9"                      # 4:3 for standard size,16:9
      slideNumberFormat: |
       <div class="progress-bar-container">
        <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
        </div>
       </div>
---
class: clear, title-slide, inverse, center, top, middle

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=FALSE}
# then load all the relevant packages
pacman::p_load(pacman, knitr, tidyverse, readxl)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

```{r xaringanExtra-clipboard, echo=FALSE}
# these allow any code snippets to be copied to the clipboard so they 
# can be pasted easily
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
  ),
  rmarkdown::html_dependency_font_awesome()
)
```
```{r xaringan-extras, echo=FALSE}
xaringanExtra::use_tile_view()

```

# `r rmarkdown::metadata$title`
----
### `r rmarkdown::metadata$author`
### `r format(Sys.time(), "%B %d, %Y")`

---

# Agenda

.pull-left[
## Part 1: Core Concepts
- Cross-validation (k-folds and v-folds)
- feature engineering
- bias-bariance trade-off
]

.pull-right[

## Part 2: R Code-Along
- MVS data
- comparing fit measures over time
]


.panelset[
.panel[.panel-name[1]

---

# How do I select a model?

One general principle is to **start with the simplest useful model** and to _build toward
more complex models as helpfuL_.

This principle applies in multiple ways:

- To choose an algorithm, start with simpler models that you can efficiently use and understand
- To carry out feature engineering, understand your predictors well by starting with a subset
- To tune an algorithm, start with a relatively simple set of tuning parameters

This isn't just for beginners or those of us in education; [most spam filters use Support Vector Machines (and used Naive Bayes until recently)](https://vas3k.com/blog/machine_learning/) due to their combination of effectiveness and efficiency "in production."

---

# Overview of regression modeling

1. Split data (into train, test, and tuning sets sets)
1. Specify recipe, model, and workflow
1. Estimate models and determine tuning parameters' values
1. Evaluate metrics and predictions

---

class: clear, center

## .font130[.center[**Thank you!**]]
<br/>
.center[<img style="border-radius: 80%;" src="img/jr-cycling" height="200px"/><br/>**Dr. Joshua Rosrenberg**<br/><mailto:jmrosenberg@utk.edu>]
