---
title: "Machine Learning Learning Lab 3"
subtitle: "Overview Presentation"
author: "**Dr. Joshua Rosenberg**"
institute: "LASER Institute"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  xaringan::moon_reader:
    css:
     - default
     - css/laser.css
     - css/laser-fonts.css
    lib_dir: libs                        # creates directory for libraries
    seal: false                          # false: custom title slide
    nature:
      highlightStyle: default         # highlighting syntax for code
      highlightLines: true               # true: enables code line highlighting 
      highlightLanguage: ["r"]           # languages to highlight
      countIncrementalSlides: false      # false: disables counting of incremental slides
      ratio: "16:9"                      # 4:3 for standard size,16:9
      slideNumberFormat: |
       <div class="progress-bar-container">
        <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
        </div>
       </div>
---
class: clear, title-slide, inverse, center, top, middle

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=FALSE}
# then load all the relevant packages
pacman::p_load(pacman, knitr, tidyverse, readxl)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

```{r xaringanExtra-clipboard, echo=FALSE}
# these allow any code snippets to be copied to the clipboard so they 
# can be pasted easily
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
  ),
  rmarkdown::html_dependency_font_awesome()
)
```
```{r xaringan-extras, echo=FALSE}
xaringanExtra::use_tile_view()

```

# `r rmarkdown::metadata$title`
----
### `r rmarkdown::metadata$author`
### `r format(Sys.time(), "%B %d, %Y")`

---

# Background

- focus: here, we wish to use uncoded data
- focus: estimating codes
- data: assistments

---

class: clear, inverse, center, middle

# Core Concepts

---

# Unsupervised methods

---

# Range of data

---

# LPA

https://data-edu.github.io/tidyLPA/index.html

---

# Computational grounded theory

---

# xx

xx

---

class: clear, inverse, center, middle

# Code Examples

---

.panelset[

.panel[.panel-name[1]

**Split data**

```{r panel-chunk-1, echo = TRUE, eval = FALSE}
library(tidyLPA)
library(dplyr)

pisaUSA15[1:100, ] %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(3)


```
]

.panel[.panel-name[2]

**Engineer features**

```{r panel-chunk-2, echo = TRUE, eval = FALSE}

```
]

.panel[.panel-name[3]

**Specify recipe, model, and workflow**
 
```{r panel-chunk-3, echo = TRUE, eval = FALSE}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(1:3, 
                      variances = c("equal", "varying"),
                      covariances = c("zero", "varying")) %>%
    compare_solutions(statistics = c("AIC", "BIC"))

```
]

.panel[.panel-name[4]

**Fit model**

```{r panel-chunk-4, echo = TRUE, eval = FALSE}
m3 <- pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    estimate_profiles(3)

get_estimates(m3)
```
]

.panel[.panel-name[5]

**Evaluate accuracy**

```{r panel-chunk-5, echo = TRUE, eval = FALSE}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    scale() %>%
    estimate_profiles(4) %>%
    plot_profiles()

get_data(m3)
get_fit(m4)
```
]
]

---

# In the remainder of this learning lab, you'll dive deeper into this model

- **Guided walkthrough**: an LPA
- **Independent practice**: your own data
- **Readings**: paper: nelson, CGT

---

class: clear, center

## .font130[.center[**Thank you!**]]
<br/>
.center[<img style="border-radius: 80%;" src="img/jr-cycling" height="200px"/><br/>**Dr. Joshua Rosrenberg**<br/><mailto:jmrosenberg@utk.edu>]
