---
title: "Using Training and Testing Data in a Workflow"
subtitle: "Conceptual Overview"
format:
  revealjs: 
    slide-number: c/t
    progress: true
    chalkboard: 
      buttons: false
    preview-links: auto
    logo: img/LASERLogoB.png
    theme: [default, css/laser.scss]
    width: 1920
    height: 1080
    margin: 0.05
    footer: <a href=https://www.go.ncsu.edu/laser-institute>go.ncsu.edu/laser-institute
---

```{r}
# then load all the relevant packages
pacman::p_load(pacman, knitr, tidyverse, readxl)
```

# Purpose and Agenda

We have some data and want to develop a prediction model. Supervised machine learning is suited to this aim. In particular, in this learning lab, we explore how we can train a computer to predict students' passing a course. We use a large data set, the Open University Learning Analytics Dataset (OULAD), focusing on student data at this point. Our model at this point is relatively simple, a generalized linear model.

## What we'll do in this presentation

- Discussion 1
- Key Concept #1: Our framework
- Key Concept #2: Training and testing data
- Discussion 2
- Introduction to the other parts of this learning lab

# Discussion 1

## Discuss!

:::{.panel-tabset}

## Getting Started

- Provide an example of supervised machine learning in the context of educational research. Discuss why this counts as machine learning.

## Digging Deeper

- How might presenting the results of a machine learning model differ from presenting those from a more traditional ("explanatory") model? 

:::

# Key Concept #1

## Our framework

::: panel-tabset

### Overview

- We want to make predictions about an outcome of interest based on predictor variables that we think are related to the outcome.
- We'll be using a widely-used data set in the learning analytics field: the [Open University Learning Analytics Dataset (OULAD)](https://analyse.kmi.open.ac.uk/open_dataset).
- The OULAD was created by learning analytics researchers at the United Kingdom-based Open University.
- It includes data from post-secondary learners participation in one of several Massive Open Online Courses (called *modules* in the OULAD).

### OULAD (1)

- Many students pass these courses, but not all do
- We have data on students' initial characteristics and their interactions in the course
- If we could develop a good prediction model, we could provide additional supports to students--and maybe move the needle on some students succeeding who might not otherwise

### OULAD (2)

We'll be focusing on three files:

- studentInfo, courses, and studentRegistration

These are joined together (see `oulad.R`) for this learning lab. You'll be doing more joining later!

```{r, message = FALSE, echo=FALSE}
students <- read_csv("data/oulad-students.csv")
students %>% head(3)
```

### LASER Frame

1. **Prepare**: Prior to analysis, we'll take a look at the context from which our data came, formulate some questions, and load R packages.
2. **Wrangle**: In the wrangling section, we will learn some basic techniques for manipulating, cleaning, transforming, and merging data.
3. **Explore**: The processes of wrangling and exploring data often go hand in hand.
4. **Model**: In this step, we carry out the analysis - here, supervised machine learning.
5. **Communicate**: Interpreting and communicating the results of our findings is the last step.

### Supervised ML Frame

1. **Split data** (Prepare)  
2. **Engineer features and write down the recipe** (Wrangle and Explore)  
3. **Specify the model and workflow** (Model)  
4. **Fit model** (Model)
5. **Evaluate accuracy** (Communicate)

**This is the fundamental process we'll follow for this and the next two learning labs focused on supervised ML**

:::

# Key Concept #2

## Training and testing

::: panel-tabset

## Algorithms

- Algorithms (or estimation procedures - or *models*) refer to the _structure_ and _process_ of estimating the _parameters_ of a model
- This definition provides a wide range of options for what kinds of algorithms we use (from simple to very complex, as we discuss in a later learning lab)
- For now, we focus on a familiar, easy to interpret algorithm (e.g., [1](https://dl.acm.org/doi/abs/10.1145/3448139.3448154?casa_token=skmk5XGbDOUAAAAA:Z0Kl4nyjpOGFA6RuFTiiLWaC_KxH1vkQ72Kr0hetXcumRdvu8tPYlCX12AgHr9aS0Fp3L-Uu0p4), also [this](https://linkinghub.elsevier.com/retrieve/pii/S0895435618310813)), _logistic regression_
- This is a linear model with a binary (*"yes"* or *"no"*) outcome
- It will be a _bad model_ to start us off!

## Predictions

- When doing supervised ML, we focus on predicting an outcome: how well we do this overall and for particular cases (more on how in the next learning lab)
- We _do not_ focus on inference or explanation (i.e., an "explanatory" model): model fit, statistical significance, effect sizes, etc.
- **This is a really key difference -- we use different metrics to evaluate what makes for a good model**

## Train vs. test
- A key concept in the context of supervised machine learning is training vs. testing data:
- Training data: Data we use to **fit** (or train, AKA estimate) a supervised machine learning **model** (AKA algorithm)
- Testing data: Data we use to see how well our model didnot used to fit the model
- By splitting out data into training and testing _sets_, we can obtain unbiased metrics for how good our model is at predicting

## How does this avoid bias?

- If we used only one data set (i.e., only training data) we could fit a model that does a _really good_ job of making predictions
- But, this model would likely be _overfit_ --- a model that is too tailored to the specific data in our training set, rather than a model that can be more _generalizable_
- The big picture, very real risk of not using training and testing data is we think we have a better model than we do
- We could fit a model that _perfectly_ predicts every outcome in our training data, but when the model sees new (i.e., different) data, it performs very poorly
- This is **essential** for supervised machine learning; if you review or see a study that only uses a single data set, be skeptical of the prediction metrics!

## Strata

- It is often valuable to conduct a _stratified_ split, based on the proportion or distribution of the dependent variables' values in the dataset
    - this ensures our training and testing data will not have a misbalance in the dependent variable

:::

# Discussion 2

::: panel-tabset

## Reflecting

- Why not use our training data to evaluate how good our model is?

## Applying

- What data or context are you interested in for your use of SML?

:::

# Introduction to the other parts of this learning lab

::: panel-tabset

## Readings

> Breiman, L. (2001). Statistical modeling: The two cultures (with comments and a rejoinder by the author). Statistical science, 16(3), 199-231.

> Estrellado, R. A., Freer, E. A., Mostipak, J., Rosenberg, J. M., & Velásquez, I. C. (2020). *Data science in education using R. Routledge* (c14), Predicting students’ final grades using machine learning methods with online course data. http://www.datascienceineducation.com/

## Case study

- Building a prediction model for students passing the class based just on student data.
- Work with peers to complete this, reading the text, following links to resources (and the reading), and then completing the required 👉 Your Turn ⤵ tasks
- A key is available, but we strongly encourage you to use it only at the end to check your work, or if you are completely stuck and have tried our recommended troubleshooting steps: https://docs.google.com/document/d/14Jc-KG3m5k1BvyKWqw7KmDD21IugU5nV5edfJkZyspY/edit

## Badge

- Involves applying what you have done through this point in the learning lab to a) extending our model and b) reflecting and planning, after which you will knit and submit your work by publishing to Posit Cloud.

:::

# *fin*

*Dr. Joshua Rosenberg* (jmrosenberg@utk.edu; https://joshuamrosenberg.com)

[General troubleshooting tips for R and RStudio](https://docs.google.com/document/d/14Jc-KG3m5k1BvyKWqw7KmDD21IugU5nV5edfJkZyspY/edit)
