---
title: 'Learning Lab 2 Case Study: Feature Engineering'
author: ""
date: "`r format(Sys.Date(),'%B %e, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
    code_folding: show
    code_download: TRUE
editor_options:
  markdown:
    wrap: 72
bibliography: lit/references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This case study is similar to the first, but it differs in three key
ways:

1.  We use a very different data set, on from *online science classes*
    that involves a variety of variables types
2.  We focus on *feature engineering*, a key step in which we prepare
    variables for inclusion in our machine learning models
3.  We use resampling to evaluate the effectiveness of the feature
    engineering steps

Feature engineering is a rich topic in machine learning research,
including in the learning analytics and educational data mining
communities.

Consider research on online learning and the work of Li et al. (2020)
and Rodriguez et al. (2021). In these two studies, digital *log-trace
data*, data generated through users' interactions with digital
technologies, was used to study elements of the theoretical frame of
*self-regulated learning* and how it related to students' achievement.
Notably, the authors took several steps to prepare the data so that it
could be validly interpreted as measures of students' self-regulated
learning. In short, we need to process the data from contexts such as
online classes to use them in analyses. Citations and links to these
papers follow.

> Li, Q., Baker, R., & Warschauer, M. (2020). Using clickstream data to
> measure, understand, and support self-regulated learning in online
> courses. The Internet and Higher Education, 45, 100727.
> <https://github.com/laser-institute/essential-readings/blob/main/machine-learning/ml-lab-2/li-et-al-2020-ihe.pdf>
>
> Rodriguez, F., Lee, H. R., Rutherford, T., Fischer, C., Potma, E., &
> Warschauer, M. (2021, April). Using clickstream data mining techniques
> to understand and support first-generation college students in an
> online chemistry course. In LAK21: 11th International Learning
> Analytics and Knowledge Conference (pp. 313-322).
> <https://github.com/laser-institute/essential-readings/blob/main/machine-learning/ml-lab-2/rodriguez-et-al-2021-lak.pdf>

The same is true here in the context of machine learning. In a different
context, the work of Gobert et al. (2013) is a great example of using
data from educational simulations. Salmeron-Majadas provides an example
of feature engineering using mouse-click data. Last, we note that there
are methods that intended to automated the process of feature
engineering (Bosch et al., 2021), though such processes are not
necessarily interpretable and they usually require some degree of
tailoring to your particular context.

> Gobert, J. D., Sao Pedro, M., Raziuddin, J., & Baker, R. S. (2013).
> From log files to assessment metrics: Measuring students' science
> inquiry skills using educational data mining. Journal of the Learning
> Sciences, 22(4), 521-563.
> <https://github.com/laser-institute/essential-readings/blob/main/machine-learning/ml-lab-3/gobert-et-al-2013-jls.pdf>

> Salmeron-Majadas, S., Baker, R. S., Santos, O. C., & Boticario, J. G.
> (2018). A machine learning approach to leverage individual keyboard
> and mouse interaction behavior from multiple users in real-world
> learning scenarios. IEEE Access, 6, 39154-39179.
> <https://ieeexplore.ieee.org/iel7/6287639/8274985/08416736.pdf>

> Bosch, N. (2021). AutoML Feature Engineering for Student Modeling
> Yields High Accuracy, but Limited Interpretability. Journal of
> Educational Data Mining, 13(2), 55-79.
> <https://github.com/laser-institute/essential-readings/blob/main/machine-learning/ml-lab-3/bosch-et-al-2021-jedm.pdf>

Our driving question for this case study is: **How much do new
predictors improve the prediction quality?**

We use a data set of many online classes to answer this question. To
answer it, we will engage in several feature engineering steps.

## Step 0: Loading and setting up

Like in the first learning lab, we'll first load several packages.

```{r, load packages}
library(tidyverse)
library(tidymodels)
```

Like in the code-along for the overview presentation, let's take a look
at the data and do some processing of it.

```{r}
d <- read_csv("data/data-to-model-no-gradebook.csv") # **edit: did this cause a problem**
d <- select(d, -time_spent) # this is another outcome, so we'll cut this here

gb <- read_csv("data/data-to-model-gradebook.csv")
```

We mentioned that this lab is premised on the need to improve on an
earlier model. Indeed, an earlier version of this model without feature
engineering achieved predictive accuracy of an RMSE of approximately 13
(see more [here](https://datascienceineducation.com/c14.html)).
Predicting students' passing (or not passing) the course with around 75%
accuracy. We think we can do better -- the aim of this learning lab is
to do just that.

As a bit more background, the online science classes we explore in this
chapter were designed and taught by instructors through a state-wide
online course provider designed to supplement---but not
replace---students' enrollment in their local school. For example,
students may have chosen to enroll in an online physics class because
one was not offered at their school. The data were originally collected
for a research study, which utilized a number of different data sources
to understand students' course-related motivation. These datasets
included:

1.  A self-report survey assessing three aspects of students' motivation
2.  Log-trace data, such as data output from the learning management
    system (LMS)
3.  Discussion board data
4.  Academic achievement data provided in a gradebook (for the first 20
    assignments students' completed)

*Data sources 1-3 are already joined together in the data frame we named
`d` above*

*Data source 4 - the gradebook data - is separate, in the data frame we
named `gb`.*

Take a look at the two data frames by running the two chunks below.

```{r}
d
```

```{r}
gb
```

You'll notice the data have different dimensions. We'll have to take
some steps to further process the gradebook data. In doing so, we'll
engineer some features. Let's take a closer look at the gradebook data.

```{r}
gb %>% 
    glimpse()
```

#### [Your Turn]{style="color: green;"} ⤵

You may also want to take a look *through* the data with the `View()`
function; try that out below (asking for help or searching the Internet
for help as needed!).

```{r, eval = FALSE}

```

Let's first consider what these variables are, focusing just on some key
variables:

-   `course_id`: an identifier for the course
-   `student_id`: an identifier for the student
-   `gradebook_item`: the name of the gradebook entry/assignment
-   `item_position`: the position of the gradebook item in the
    gradebook; differs between students
-   `grade_category`: `Hw` (homework), `Qz` (quiz or test), or `NA` (not
    classified)
-   `points_earned`: the number of points student earned
-   `points_possible`: the number of points possible to earn

What are some *features* we could create based on these variables? And
how might we create them?

#### [Your Turn]{style="color: green;"} ⤵

Add a few ideas below before proceeding:

-   

-   

```{r}
gb
```

Let's get to feature engineering. First, we'll have to *group* our data
by course and student ID.

```{r}
gb <- gb %>% 
    group_by(course_id, student_id)
```

#### [Your Turn]{style="color: green;"} ⤵

Next, let's create a variable with the percent of points earned (points
earned divided by points **possible**). To do so, add to the `mutate()`
function below. Create a new variable called `percent_earned`. You can
read more about mutate [here](https://r4ds.had.co.nz/transform.html)

```{r, warning = FALSE}
gb <- gb %>% 
    mutate()
```

Finally, let's create three features from the gradebook data:

-   The overall percent of points earned (*across* all assignments)
-   The variability (in standard deviation units) in the percent earned
    (*between* assignments)
-   The number of assignments for which students earned 100% of the
    possible points

You can probably imagine others; you're welcome to explore adding those,
too.

We'll use *summarize* to do this, as below:

```{r}
gb <- gb %>% 
    summarize(overall_percent_earned = sum(points_earned, na.rm = TRUE) / sum(points_possible, na.rm = TRUE),
              variability_percent_earned = sd(percent_earned, na.rm = TRUE),
              n_with_100_pct = sum(percent_earned == 1, na.rm = TRUE)) %>% 
    select(student_id, course_id, overall_percent_earned, variability_percent_earned, n_with_100_pct) # selecting just the variables we'll use
```

We have one last step before we can get to modeling (`gb`) - joining
this data with all of the other data (`d`).

```{r}
d <- d %>% 
    left_join(gb)
```

#### [Your Turn]{style="color: green;"} ⤵

Let's talk a look at the joined data to make sure everything is looking
as we intend it to. Inspect the data using the code chunk below:

```{r}

```

## Step 1. Split data

Next, we'll split the data, just like before. We'll set the seed again
to ensure that we obtain the same results (when running the analysis
again and between analysts at the LASER Institute). We use an 80% split
again; how will you "spend" your data? You can change this number if you
wish, but consider how much data you have to "spend" for both training
and testing.

```{r}
set.seed(20220712)

train_test_split <- initial_split(d, prop = .80)

data_train <- training(train_test_split)
```

Here's a key difference! Pay careful attention to this next line of
code, which sets the groundwork for *k*-folds cross-validation. Note
that in the function below (run `?vfold_cv` to see more), the letter *v*
is used instead of *k*, though they share a meaning, as the
documentation notes).

```{r}
kfcv <- vfold_cv(data_train) # this differentiates this from what we did before
# before, we simple used data_train to fit our model
kfcv
```

#### [Your Turn]{style="color: green;"} ⤵

Above, we split the data into 10 different folds. Change the number of
folds from 10 to 20 by changing the value of v; 10 is simply the
default. For help, run `?vfold_cv` to get a hint.

```{r}
kfcv <- vfold_cv(data_train) # this differentiates this from what we did before
# before, we simple used data_train to fit our model
kfcv
```

## Step 2: Engineer features and write down the recipe

Here, we'll carry out several feature engineering steps.

#### [Your Turn]{style="color: green;"} ⤵

Read about [possible steps](https://www.tmwr.org/recipes.html) and see
more about how the following five feature engineering steps below work.
Like in the first learning lab, this is the step in which we set the
recipe.

-   `step_normalize(all_numeric_predictors())`
-   `step_nzv(all_predictors())`
-   `step_novel(all_nominal_predictors())`
-   `step_dummy(all_nominal_predictors())`
-   `step_impute_knn(all_predictors(), all_outcomes())`

```{r,}
recipe(final_grade ~ ., data = data_train) %>% # **edit**: change t final grade
    update_role(student_id, course_id, section, new_role = "ID variable") %>% # this can be any string; for ID variables
    step_normalize(all_numeric_predictors()) %>% # standardizes numeric variables
    step_nzv(all_predictors()) %>% # remove predictors with a "near-zero variance"
    step_dummy(all_nominal_predictors()) %>%  # dummy code all factor variables
    step_novel(all_nominal_predictors()) %>% # add a musing label for factors
    step_impute_knn(all_predictors()) # impute missing data for all predictor variables
```

## Step 3: Specify the model and workflow

Next, we specify the model and workflow, using the same engine *but a
different engine and mode*, here, regression for a *continuous outcome*.
Specifically, we use:

-   using the `linear_reg()` function to set the *model*
-   using `set_engine("glm")` to set the *engine*
-   finally, using `set_mode("regression"))`

```{r}
# specify model
my_mod <-
    linear_reg() %>% 
    set_engine("glm") %>%
    set_mode("regression")
```

Last, we'll put the pieces together - the model and recipe - in a
workflow.

```{r}
# specify workflow
my_wf <-
    workflow() %>%
    add_model(my_mod) %>% 
    add_recipe(my_rec)
```

## Step 4: Fit model

Note that here we use the `kfcv` data. We'll run that in the next chunk.

We can ignore the warnings and messages we see.

```{r, warning = FALSE}
fitted_model_resamples <- fit_resamples(my_wf, resamples = kfcv,
                              control = control_grid(save_pred = TRUE)) # this allows us to inspect the predictions
```

## Step 5: Interpret accuracy

What did we get as output? Let's take a look at the metrics. This is
critical to understanding how and why we use k-folds cross validation.
Each of the rows below represents the accuracy (in the `.estimate`
column) for each of the 20 folds that we used to train our model; our
model was fit 20 times, and accuracy was calculated separately for each
of these times. Next, we'll summarize these.

Recall our definition of the Root Mean Squared Error (RMSE) - it is the
*square root* of the mean of the squared error, or difference between
the predicted and known *y* variables (here, students' final grade).
Since this is the square root of a statistic that is squared, its
interpretation can be considerably simplified: **RMSE can be interpreted
as the average error, or difference between the predicted and known *y*
variables (here, students' final grade)**. This, along with the Mean
Squared Error (MSE), are the most common metrics of predictive accuracy
for a numeric outcome such as students' final grade. See more about fit
metrics for numeric/continuous outcomes (those utilized in a
*regression* mode)
[here](https://bradleyboehmke.github.io/HOML/process.html#regression-models).
The goal is to minimize both the RMSE and MSE.

Note that the common R-squared measure (`rsq` in the output) can also be
interpreted. Though helpful descriptively, it has less useful as a
measure of the predictive effectiveness of a trained model, and it
should generally not be used to select between competing model
specifications.

```{r}
fitted_model_resamples %>% # **edit: this needs to be fitted_model_resamples** 
    unnest(.metrics) %>% 
    filter(.metric == "rmse") # we also get another metric, the RSQ; we focus just on RMSE for nwo
```

Running the code below calculates the *mean* of the metrics we inspected
in the previous chunk. Focus on the **mean** variable for the accuracy
metric. This can be interpreted in the precise same was as our accuracy
measure we calculated in learning lab 1 - this is the percentage of
students the model correctly classified as passing or not passing the
course.

```{r}
# fit stats
fitted_model_resamples %>%
    collect_metrics()
```

We can imagine trying out many different sets of features (engineered in
different ways). So long as we evaluate the accuracy using the
resampling method used above, we can repeat this process as needed.
Then, we can carry out a process like that in the first learning lab -
fitting the model not using the different *folds* obtained through the
`kfcv` function, but rather using the **entire training data set**.

```{r, warning = FALSE}
fitted_model <- fit(my_wf, data_train)
```

Then, we can use the model to predict students passing (or not passing)
using our testing data *that we have not used for any purpose until this
point* --- and interpret that model. This output is suggestive to us of
how the model would perform on new data, as this testing data set has
not been used to make any decisions about the feature engineering.

```{r}
final_fit <- last_fit(fitted_model, train_test_split)
```

```{r}
collect_metrics(final_fit)
```

Last, we can plot the predicted versus known *y* variables to gain a
graphical sense for how the model performed:

```{r}
collect_predictions(final_fit) %>% 
    ggplot(aes(x = .pred, y = final_grade)) +
    geom_point()
```

Consider making a modification to the above plot (small or large) using
ggplot2.

### 🧶 Knit & Check ✅

Congratulations - you've completed this case study! Consider moving on
to the badge activity next.
