---
title: "Machine Learning Learning Lab 1: Foundations"
subtitle: "Overview Presentation"
author: "**Dr. Joshua Rosenberg**"
institute: "LASER Institute"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  xaringan::moon_reader:
    css:
     - default
     - css/laser.css
     - css/laser-fonts.css
    lib_dir: libs                        # creates directory for libraries
    seal: false                          # false: custom title slide
    nature:
      highlightStyle: default         # highlighting syntax for code
      highlightLines: true               # true: enables code line highlighting 
      highlightLanguage: ["r"]           # languages to highlight
      countIncrementalSlides: false      # false: disables counting of incremental slides
      ratio: "16:9"                      # 4:3 for standard size,16:9
      slideNumberFormat: |
       <div class="progress-bar-container">
        <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
        </div>
       </div>
---

class: clear, title-slide, inverse, center, top, middle

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
```

```{r, echo=FALSE}
# then load all the relevant packages
pacman::p_load(pacman, knitr, tidyverse, readxl)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

```{r xaringanExtra-clipboard, echo=FALSE}

```

# `r rmarkdown::metadata$title`
----
### `r rmarkdown::metadata$author`
### `r format(Sys.time(), "%B %d, %Y")`


---

# Code-along

.panelset[

.panel[.panel-name[0]

**Loading, setting up: create a .R file in /lab-1 and run this code**

```{r, eval = FALSE, echo = TRUE}
library(tidyverse)
library(tidymodels)

starwars_recoded <- starwars %>% # built-in data available just by typing
    mutate(species_human = ifelse(species == "Human", "Human", "Not human")) # recoding species to create a categorical variable

starwars_recoded %>% 
    count(species) # how many humans are there?
```

]

.panel[.panel-name[1]

**Split data**

```{r, echo = TRUE, eval = FALSE}
train_test_split <- initial_split(starwars_recoded, prop = .80)

data_train <- training(train_test_split)
data_test <- testing(train_test_split)
```

]

.panel[.panel-name[2]

**Engineer features**

```{r, echo = TRUE, eval = FALSE}
# predicting humans based on the independent effects of height and mass
my_rec <- recipe(species_human ~ height + mass, data = data_train)
```

]

.panel[.panel-name[3]

**Specify recipe, model, and workflow**

```{r, echo = TRUE, eval = FALSE}
# specify model
my_mod <- logistic_reg() %>% 
    set_engine("glm") %>%
    set_mode("classification")

# specify workflow
my_wf <- workflow() %>%
    add_model(my_mod) %>% 
    add_recipe(my_rec)
```

]

.panel[.panel-name[4]

**Fit model**

```{r, echo = TRUE, eval = FALSE}
fitted_model <- fit(my_wf, data = data_train)

final_fit <- last_fit(fitted_model, train_test_split)
```

]

.panel[.panel-name[5]

**Evaluate accuracy**

```{r, echo = TRUE, eval = FALSE}
final_fit %>%
    collect_metrics()
```

]


]

---

# Discussion 2

.panelset[

.panel[.panel-name[Reflecting]

- When might you use SML?

]

.panel[.panel-name[Applying]

- How could you use SML?

]

]

---

# Introduction to the other parts of this learning lab

.panelset[

.panel[.panel-name[Readings]

Breiman, L. (2001). Statistical modeling: The two cultures (with comments and a rejoinder by the author). *Statistical Science, 16*(3), 199-231. https://projecteuclid.org/journals/statistical-science/volume-16/issue-3/Statistical-Modeling--The-Two-Cultures-with-comments-and-a/10.1214/ss/1009213726.pdf

Estrellado, R. A., Freer, E. A., Mostipak, J., Rosenberg, J. M., & Velásquez, I. C. (2020). *Data science in education using R. Routledge* (c14), Predicting students’ final grades using machine learning methods with online course data. http://www.datascienceineducation.com/

]

.panel[.panel-name[Case Study]

- Building a prediction model for students passing the class---based just on student data.
- Work with peers to complete this, reading the text, following links to resources (and the reading), and then completing the required 👉 Your Turn ⤵ tasks
- A key is available, but we strongly encourage you to use it only at the end to check your work, or if you are completely stuck and have tried our recommended troubleshooting steps: https://docs.google.com/document/d/14Jc-KG3m5k1BvyKWqw7KmDD21IugU5nV5edfJkZyspY/edit

]

.panel[.panel-name[Badge]

- Involves applying what you have done through this point in the learning lab to a) extending our model and b) reflecting and planning, after which you will knit and submit your work by publishing to Posit Cloud.

]
]

---

# Troubleshooting and setup

- Please view [this Google doc](https://docs.google.com/document/d/14Jc-KG3m5k1BvyKWqw7KmDD21IugU5nV5edfJkZyspY/edit)
- Change default settings: Tools -> Global Settings -> Workspace -> DO NOT restore workspace 
- Change default settings: Tools -> Global Settings -> Workspace -> Save workspace to .RData NEVER
- Session -> Restart R and Clear Output is a good place to start

---

# *fin*

We hope to see you in the ML topic area learning labs!

- *Dr. Joshua Rosenberg* (jmrosenberg@utk.edu; https://joshuamrosenberg.com)

[These slides (Introductory Presentation)](https://laser-institute.github.io/machine-learning/introductory-presentation.html#1)

[General troubleshooting tips for R and RStudio](https://docs.google.com/document/d/14Jc-KG3m5k1BvyKWqw7KmDD21IugU5nV5edfJkZyspY/edit)