---
title: 'Learning Lab 3 Case Study'
author: ""
date: "`r format(Sys.Date(),'%B %e, kf%Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
    code_folding: show
    code_download: TRUE
editor_options:
  markdown:
    wrap: 72
bibliography: lit/references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

## 1. PREPARE

After interpreting our last model, it is easy to think we can do a
little better. But, how? In this learning lab, we will answer this
question will building a better model.

Feature engineering is a rich topic in machine learning research,
including in the learning analytics and educational data mining
communities.

Consider research on online learning and the work of Rodriguez et al.
(2021). In these two studies, *digital trace data*, data generated
through users' interactions with digital technologies. Optionally,
review this paper -- specifically how they processed the "clickstream"
data. As this paper illustrates, there is not one way to use such data.

> Rodriguez, F., Lee, H. R., Rutherford, T., Fischer, C., Potma, E., &
> Warschauer, M. (2021, April). Using clickstream data mining techniques
> to understand and support first-generation college students in an
> online chemistry course. In *LAK21: 11th International Learning
> Analytics and Knowledge Conference* (pp. 313-322).
> <https://github.com/laser-institute/essential-readings/blob/main/machine-learning/ml-lab-2/rodriguez-et-al-2021-lak.pdf>

Notably, the authors took several steps to prepare the data so that it
could be validly interpreted. The same is true here in the context of
machine learning. In a different context, the work of Gobert et al.
(2013) is a great example of using data from educational simulations.
Optionally review this paper, too, focused on their use of a technique
they called *replay tagging* to conduct feature engineering.

> Gobert, J. D., Sao Pedro, M., Raziuddin, J., & Baker, R. S. (2013).
> From log files to assessment metrics: Measuring students' science
> inquiry skills using educational data mining. *Journal of the Learning
> Sciences, 22*(4), 521-563.
> <https://github.com/laser-institute/essential-readings/blob/main/machine-learning/ml-lab-3/gobert-et-al-2013-jls.pdf>

Even after feature engineering, machine learning approaches can often
(but not always) be improved by choosing a more sophisticated model
type. Note how we used a regression model in the first two case studies;
here, we explore a considerably more sophisticated model, a random
forest. Feature engineering and choosing a more sophisticated model adds
some complexity to the modeling. As we have discussed, it is easy to
bias our results if we repeatedly check the performance of *different
model fits* with the same test data. Cross-validation is one commonly
used solution for this problem.

Our driving question is: **How much can we improve our model?** Looking
back to our predictive model from LL 2, we can see that our accuracy was
okay. Can we improve on that? Let's dive in!

## 2. WRANGLE

## Step 0: Loading and setting up

First, let's load the packages we'll use---the familiar {tidyverse} and
several others focused on modeling.

#### **👉 Your Turn** **⤵**

Please add to the chunk below code to load three packages we've used
before -- tidyverse, janitor, and tidymodels.

Please install and add two additional packages: {ranger}, for one
implementation of random forest modeling, and {vip}, for variable
importance metrics. Load these with the library function

```{r}
library(tidyverse)
library(janitor)
library(tidymodels)


```

Next, we'll load a new file --- one with *interactions* (or log-trace)
data. Please find that in the `data` folder for this lab, read that in
and assign it the name `interactions`.

```{r}

```

We have to do the same processing we did in the second learning lab, to
obtain cut-off dates. As a reminder, the purpose of this is to train the
model on data from the first one-third of the class, with the reasoning
being this is a good time to intervene--far enough into the class to
make an appreciable impact, but not too late to have a limited chance of
being able to change students' trajectory in the class.

We'll repeat the procedure we carried out with the assessments data ---
calculating a cut-off for each class and then filtering the data based
upon this cut-off. But, since we've already done this for the assessment
data, to allow us to focus more on some new feature engineering steps in
this learning lab, we are providing you with the already-filtered
interactions data. *Note*: you can find the code used to carry out these
steps in the Lab 3 folder in `process-interactions-data.R`.

Please load `oulad-interactions-filtered.csv` into R, assigning the
resulting data frame the name `interactions`.

#### **👉 Your Turn** **⤵**

```{r}



```

In the OULAD documentation, this is called the VLE (virtual learning
environment) data source. Please review the description of the variables
in the *studentVLE* and *VLE* sources (which are joined together for
this learning lab)
[here](https://analyse.kmi.open.ac.uk/open_dataset#description). Then,
read in the `oulad-interactions.csv` file.

Woah, even filtered, this is a large file! Let's start to explore it a
bit. Please take three steps to explore the data, as follows.

## 3. EXPLORE

#### **👉 Your Turn** **⤵**

*First*, `count()` the `activity_type` variable and *sort* the resulting
output by frequency.

```{r}

```

What does this tell you? Consulting the codebook and your output, please
add at least two notes on what you are noticing:

-   ADD YOUR RESPONSE HERE

#### **👉 Your Turn** **⤵**

*Second*, please create a histogram of the `date` variable.

```{r}

```

What does this tell you? Add one or more notes:

-   ADD YOUR RESPONSE HERE

#### **👉 Your Turn** **⤵**

*Third*, please conduct one other data exploration step of your
choosing. Options include creating simple graphgs or calculating
descriptive, summary statistics.

```{r}

```

We're ready to proceed to engineering some features with the
interactions data.

First, let's join together the `interactions` and the
`code_module_dates` data frames.

We'll use the same code to create the `code_module_dates` we used in
LL2.

```{r}
assessments <- read_csv("data/oulad-assessments.csv")

code_module_dates <- assessments %>% 
    group_by(code_module, code_presentation) %>% 
    summarize(quantile_cutoff_date = quantile(date, probs = .25, na.rm = TRUE))
```

This is so we have the cut-offs (one-third through the semester), which
will help us to only use data from that period in our model when we
filter the data based upon those dates.

```{r}
interactions_joined <- interactions %>% 
  left_join(code_module_dates) # join the data based on course_module and course_presentation

interactions_filtered <- interactions_joined %>% 
  filter(date < quantile_cutoff_date) # filter the data so only assignments before the cutoff date are included
```

Then, let's engineer several features. For the present time, we'll focus
on the `sum_click` variable, which tells us how many times students
clicked on a resource for a given date. We'll use the
`interactions_filtered` data set we just created. Let's start simple:
please type the name of that data set (`interactions_filtered`) in the
code chunk below.

#### **👉 Your Turn** **⤵**

```{r}

```

How can we create a feature with `sum_click`? Think back to our
discussion in the presentation; we have *many* options for working with
such time series data. Perhaps the most simple is to count the clicks.
Please summarize the number of clicks for each student (specific to a
single course). This means you will need to group your data by
`id_student`, `code_module`, and `code_presentation`, and then create a
summary variable. Assign the resulting output the name
`interactions_summarized`. You may find the documentation for
`summarize()` to be helpful. That is available
[here](https://dplyr.tidyverse.org/reference/summarise.html). [This
chapter](https://r4ds.had.co.nz/transform.html) is also likely helpful
(note that `summarise()` and `summarize()` are different spellings for
the same function).

#### **👉 Your Turn** **⤵**

```{r}

```

#### **👉 Your Turn** **⤵**

How many times did students click? Let's create a histogram to see.
Please use {ggplot} and `geom_histogram()` to visualize the distribution
of the `sum_clicks` variable you just created. Turn to [the
documentation](https://ggplot2.tidyverse.org/reference/geom_histogram.html)
if you need a pointer!

```{r}

```

#### [Your Turn]{style="color: green;"} ⤵

How many times did students click? Let's create a histogram to see.
Please use {ggplot} and `geom_histogram()` to visualize the distribution
of the `sum_clicks` variable you just created. Turn to [the
documentation](https://ggplot2.tidyverse.org/reference/geom_histogram.html)
if you need a pointer!

```{r}

```

This is a good start - we've created our first feature based upon the
log data, `sum_clicks`! What are some other features we can add? An
affordance of using the `summarize()` function in R is we can create
multiple summary statistics at once. Let's also calculate the standard
deviation of the number of clicks as well as the mean. Please copy the
code you wrote above into the code chunk below and then add these two
additional summary statistics.

#### **👉 Your Turn** **⤵**

```{r}

```

We'll take one last step here -- creating individual *slopes* of
students' clicks over time. This code is a bit more involved, and so is
annotated for you below; feel free to modify and re-use this.

**Note: this may take up to a minute or two to run! Please feel free to
relax!**

```{r}
interactions_slopes <- interactions_filtered %>% # start with our interactions data
  group_by(id_student, code_module, code_presentation) %>% # group by student and course so a slope is calculated for each student by course combination 
  nest() %>% # this tells R that we want to create a smaller data frame, wherein each students' data is assigned its own cell; it makes the next step possible
  mutate(model = map(data, ~lm(sum_click ~ 1 + date + I(date^2), data = .x) %>% 
                       tidy)) %>% # now, given that we have a smaller data frame with each student by course combination in its own cell, we can use mutate to fit a linear model for each students by course data set; the linear model fits a linear and a quadratic term
  unnest(model) # this returns the data to its original (non-nested structure), retaining the mutated model output
```

Please rename the resulting intercept, date, and data squared variables
to be renamed: `intercept`, `date`, and `date_2`, respectively.

#### **👉 Your Turn** **⤵**

```{r}

```

After running this code, we will have intercepts, linear slopes, and
quadratic terms for each students' clickstream pattern over the
semester.

Let's join together several files. First, let's join all our features
into a single file. Please use `left_join()` to join
`interactions_filtered` and `interactions_slopes`, assigning the
resulting output the name `all_features`.

#### **👉 Your Turn** **⤵**

```{r}

```

Just one last step! Let's join together *all* of the data we'll use for
our modeling: `students_and_assessments` and `all_features`. Use
`left_join()` once more, assigning the resulting output the name
`students_assessments_and_interactions`.

#### **👉 Your Turn** **⤵**

```{r}

```

One more small step. Let's ensure that our outcome variable -- `pass` --
is a factor. It was, but when saving to and reading from a CSV, it lots
its factor characteristics, becoming instead a character string. It is
necessary for the outcome of a classification model (like the one we are
using) to be a factor. Please use `mutate()` to do just that.

#### **👉 Your Turn** **⤵**

```{r}



```

We're now ready to model!

## 4. MODEL

### Step 1. Split data

We'll follow the same steps we followed in learning labs #1 and #2,
here. One difference - we'll use `students_assessments_and_interactions`
instead of the data frame we used in those learning labs. Please port
over the code you used in those learning labs here, changing the name of
the data frame to the one we are now using.

We discuss this first step minimally as we have now carried out a step
very similar to this in LL1 and LL2; return to the case study for those
(especially LL1) for more on data splitting.

#### **👉 Your Turn** **⤵**

```{r}

```

There is a key difference that is next. In this step. we'll further
process `data_train`, creating different subsets of the data, or folds
of the data, that we can use to fit our model multiple times.

```{r}
vfcv <- vfold_cv(data_train) # this differentiates this from what we did before
# before, we simple used data_train to fit our model
vfcv
```

How does this work? `data_train` is sampled as many time as we se \####
[Your Turn]{style="color: green;"} ⤵

Above, we split the data into 10 different folds. Change the number of
folds from 10 to 20 by changing the value of v; 10 is simply the
default---not always the best one! For help, run `?vfold_cv` to get a
hint.

```{r}
kfcv <- vfold_cv(data_train, v = 20) # this differentiates this from what we did before
# before, we simple used data_train to fit our model
kfcv
```

### Step 2: Engineer features and write down the recipe

Here, we'll carry out several feature engineering steps.

Read about [possible steps](https://www.tmwr.org/recipes.html) and see
more about how the following five feature engineering steps below work.
Like in the first learning lab, this is the step in which we set the
recipe.

-   [`step_normalize()`](https://recipes.tidymodels.org/reference/step_normalize.html):
    normalizes numeric data to have a standard deviation of one and a
    mean of zero
-   [`step_dummy()`](https://recipes.tidymodels.org/reference/step_dummy.html):
    convert nominal data (e.g. character or factors) into one or more
    numeric binary model terms for the levels of the original data.
-   [`step_impute_knn()`](https://recipes.tidymodels.org/reference/step_impute_knn.html):
    impute missing alues using the nearest neighbors method

We will use all three of these. Please note that there are two ways to
use each of these. One of them is by *specifying the specific variable*
you want to use for the feature engineering step. For example, below, we
would normalize a hypothetical variable, one called
`minutes_of_videos_watched`:

```{r, eval = FALSE}
step_normalize(minutes_of_videos_watched)
```

Another way is to specify *for which types of variables* the feature
engineering step applies. You can see all of the available types
[here](https://recipes.tidymodels.org/reference/has_role.html). For
example, the code below would normalize all numeric variables:

```{r, eval = FALSE}
step_normalize(all_numeric_predictors())
```

#### **👉 Your Turn** **⤵**

Let's turn to our recipe. Please *add to* the code below to add the new
variables---features---we created. We've started you off with \_\_\_
code below (where you can add the new variables -- be sure to add all of
them!).

```{r}
my_rec <- recipe(pass ~ disability +
                   date_registration + 
                   gender +
                   code_module +
                   mean_weighted_score +
                   ___ + ___ + ___,
                 data = data_train) 
```

#### **👉 Your Turn** **⤵**

Then, please *add to* the following code by completing three steps,
building on the code that is started below:

1.  Please normalize all numeric predictors.
2.  Please dummy code the following three variables: `disability`,
    `gender`, and `code_module`.
3.  Please impute missing values using the nearest neighbors method for
    the `sum_clicks` and `date_registration` variables.

```{r}

```

### Step 3: Specify the model and workflow

Next, we specify the model and workflow, using the same engine *but a
different engine and mode*, here, regression for a *continuous outcome*.
Specifically, we are:

-   using the `random_forest()` function to set the *model* as a random
    forest
-   using `set_engine("ranger", importance = "impurity")` to set the
    *engine* as that provided for random forests through the {ranger}
    package; we also add the `importance = "impurity"` line to be able
    to interpret a particular variable importance metric specific to
    random forest models
-   finally, using `set_mode("classification"))` as we are again
    predicting categories (transactional and substantive conversations
    taking place through #NGSSchat)

```{r panel-chunk-3, echo = TRUE, eval = TRUE}
# specify model
my_mod <-
  rand_forest() %>% # specify the type of model we are fitting, a random forest
  set_engine("ranger", importance = "impurity") %>% # using a variable importance metric specific to this random forest engine
  set_mode("classification") # same as before

# specify workflow
my_wf <-
  workflow() %>% # create a workflow
  add_model(my_mod) %>% # add the model we wrote above
  add_recipe(my_rec) # add our recipe we wrote above
```

### Step 4: Fit model

Note that here we use the `kfcv` data and a different function -
`fit_resamples()`, instead of `fit()`. This function is required when we
have multiple folds of our sample, as we created using the vfcv function
above. See more about this function
[here](https://tune.tidymodels.org/reference/fit_resamples.html).

```{r}
class_metrics <- metric_set(accuracy, sensitivity, specificity, ppv, npv, kap) # specify the same metrics as earlier
fitted_model_resamples <- fit_resamples(my_wf, resamples = vfcv, metrics = class_metrics) # specify the workflow, resampled data, and our metrics
```

Note that you have fit as many models as the value for `v` that you
specified earlier. So, this may take some time. Take a walk, grab a
snack, or make a cup of tea!

Then, we can use the same `collect_metrics()` function we have used to
inspect the predictive strength of the model. Please do that for
`fitted_model_resamples`.

#### **👉 Your Turn** **⤵**

```{r}

```

We noted earlier that we may introduce bias when we evaluate different
models against the same training data. The benefit of using the multiple
folds - and fitting multiple models - is we can make changes to our
features or model. Thus, if we decide to add a new feature, we can run
the above steps, without concern about biasing our interpretation of how
accurate our model is.

That is, we can be unconcerned about inadvertently introducing bias
until we reach the last step - fitting our last model. We can do that
using the familiar `fit()` function, followed by `last_fit`. It is
important to not make further changes after this point.

```{r}
fitted_model <- fit(my_wf, data_train)
final_fit <- last_fit(fitted_model, train_test_split, metrics = class_metrics)
```

### Step 5: Interpret accuracy

fa Finally, collect the metrics for our final fit. *These* are the
values that you should report as final. \#### [Your
Turn]{style="color: green;"} ⤵

```{r}
final_fit %>% 
  collect_metrics()
```

We can see that `final_fit` is for a single fit: a random forest with
the best performing tuning parameters trained with the *entire* training
set of data to predict the values in our (otherwise not used/"spent")
testing set of data.

#### **👉 Your Turn** **⤵**

```{r}


```

## 5. COMMUNICATE

Another benefit of a random forest is we can interpret variable
importance metrics. Do that here with the following code.

```{r}
final_fit %>% 
  pluck(".workflow", 1) %>%   
  extract_fit_parsnip() %>% 
  vip(num_features = 10)
```

Please add two or more notes on what you notice about which variables
(features) are important, focused on what you would say to someone in
your audience about what the important variables in your model were.

-   ADD YOUR RESPONSE HERE

-   ADD YOUR RESPONSE HERE

### 🧶 Knit & Check ✅

Congratulations - you've completed this case study! Consider moving on
to the badge activity next.
