---
title: "Machine Learning Learning Lab 4: Unsupervised Methods"
subtitle: "Overview Presentation"
author: "**Dr. Joshua Rosenberg**"
institute: "LASER Institute"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  xaringan::moon_reader:
    css:
     - default
     - css/laser.css
     - css/laser-fonts.css
    lib_dir: libs                        # creates directory for libraries
    seal: false                          # false: custom title slide
    nature:
      highlightStyle: default         # highlighting syntax for code
      highlightLines: true               # true: enables code line highlighting 
      highlightLanguage: ["r"]           # languages to highlight
      countIncrementalSlides: false      # false: disables counting of incremental slides
      ratio: "16:9"                      # 4:3 for standard size,16:9
      slideNumberFormat: |
       <div class="progress-bar-container">
        <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
        </div>
       </div>
---

class: clear, title-slide, inverse, center, top, middle

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
```

```{r, echo=FALSE}
# then load all the relevant packages
pacman::p_load(pacman, knitr, tidyverse, readxl)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

```{r xaringanExtra-clipboard, echo=FALSE}

```

# `r rmarkdown::metadata$title`
----
### `r rmarkdown::metadata$author`
### `r format(Sys.time(), "%B %d, %Y")`

---


# Purpose and Agenda

We have some data and want to develop a prediction model. Supervised machine learning is suited to this aim. In particular, in this learning lab, we explore how we can train a computer to predict students' withdrawal from a course. We use a large data set, the Open University Learning Analytics Dataset (OULAD), focusing on student data at this point. Our model at this point is relatively simple, a generalized linear model.

## What we'll do in this presentation

- Discussion 1
- Key Concepts
- Code-along
- Discussion 2
- Introduction to the other parts of this learning lab

---

# Discussion 1

.panelset[

.panel[.panel-name[Discussion]

- 
- 
-
    
]

.panel[.panel-name[Conceptual Overview]

- 
- 
-

]

.panel[.panel-name[Coding Walkthrough]

- 
- 
- 

]
]

---

# Key Concepts

.panelset[

.panel[.panel-name[Discussion]

- 
- 
-
    
]

.panel[.panel-name[Conceptual Overview]

- 
- 
-

]

.panel[.panel-name[Coding Walkthrough]

- 
- 
- 

]
]

---

# Code-along

.panelset[

.panel[.panel-name[Discussion]

- 
- 
-
    
]

.panel[.panel-name[Conceptual Overview]

- 
- 
-

]

.panel[.panel-name[Coding Walkthrough]

- 
- 
- 

]
]

---

# Discussion 2

.panelset[

.panel[.panel-name[Discussion]

- 
- 
-
    
]

.panel[.panel-name[Conceptual Overview]

- 
- 
-

]

.panel[.panel-name[Coding Walkthrough]

- 
- 
- 

]
]

---

# Introduction to the other parts of this learning lab

.panelset[

.panel[.panel-name[Discussion]

- 
- 
-
    
]

.panel[.panel-name[Conceptual Overview]

- 
- 
-

]

.panel[.panel-name[Coding Walkthrough]

- 
- 
- 

]
]

